Ship-based lidar evaluation of Southern Ocean clouds in the storm-resolving general circulation model ICON, and the ERA5 and MERRA-2 reanalyses

### 2024-12-11 manuscript_agu.tex
Addressing comments

Here, we conventionally refer to the SO as ocean regions south of 40°S, low-latitude SO as 40--55°S and high-latitude SO as south of 55°S. The reason for this dividing latitude is to split the SO into about two equal zones, as well as the results by \citeA{schuddeboom2021} (Fig. 2b), showing a contrast in CMIP model radiation biases, \citeA{schuddeboom2019} (Fig. 2) and \citeA{kuma2020} (Fig. 3) showing contrasting radiation biases in the Hadley Centre Global Environmental Model, and \citeA{cesana2022} showing contrasting cloud biases due to the 0°C isotherm reaching the surface at 55°S. The findings of \citeA{niu2024}, however, support a different dividing line of 62°S based on cloud condensation nuclei concentration.

@@

Together, we analyzed data from 31 voyages of RV \emph{Polarstern}, the resupply vessel (RSV) \emph{Aurora Australis}, RV \emph{Tangaroa}, RV \emph{Nathaniel B. Palmer}, Her (now His) Majesty's New Zealand Ship (HMNZS) \emph{Wellington} and one sub-Antarctic station (Macquarie Island) in the SO south of 40°S between 2010 and 2021. Fig. \ref{fig:map} shows a map of the campaigns, Table \ref{tab:voyages} lists the campaigns, and Table \ref{tab:voyage-references} lists references where available. Altogether, the analyzed dataset comprised 2421 days of data south of 40°S, but the availability of ceilometer data was slightly shorter due to gaps in measurements.

@@

The Vaisala CL51 and CT25K (photos in Fig. \ref{fig:map}d, e) are ceilometers operating at near-infrared wavelengths of 910 nm and 905 nm, respectively. The CL51 can also be configured to emulate the Vaisala CL31. The maximum range is 15.4 km (CL51), 7.7 km (CL31 emulation mode with 5 m vertical resolution), and 7.5 km (CT25K). The vertical resolution is 10 m (5 m configurable) in CL51 and 30 m in CT25K observations. The sampling (temporal) resolution is configurable, and in our datasets, it is approximately 6 s for CL51 on AA15‐16, 16 s for CT25K on MARCUS and MICRE, 36 s for CL51 on RV \emph{Polarstern}, and about 2.37 s for CL51 with CL31 emulation on TAN1502. The wavelengths of 905 and 910 nm are affected by water vapor absorption of about 20\% in the mid-latitudes \cite{wiegner2015,wiegner2019}, with 910 nm affected more strongly, but we do not expect this to be a significant issue as explained in \citeA{kuma2021}. The instrument data files containing raw uncalibrated backscatter were first converted to Network Common Data Form (NetCDF) with cl2nc (\url{https://github.com/peterkuma/cl2nc}) and then processed with the ALCF (Section \ref{sec:alcf}) to produce absolutely calibrated attenuated volume backscattering coefficient (AVBC), cloud mask, cloud occurrence by height, and the total cloud fraction. Because the CT25K uses a very similar wavelength to CL51, equivalent calculations as for CL51 were done assuming a wavelength of 910 nm. The Vaisala CL51 and CT25K instruments were used on most of the voyages and stations analyzed here. Fig. \ref{fig:example}a shows an example of AVBC derived from the CL51 instrument data.

@@

The Lufft CHM 15k (photo in Fig. \ref{fig:map}c) is a ceilometer operating at a near-infrared wavelength of 1064 nm. The maximum range is 15.4 km; the vertical resolution is 5 m in the near range (up to 150 m) and 15 m above; the sampling (temporal) resolution is 2 s; and the number of vertical levels is 1024. NetCDF files containing uncalibrated backscatter produced by the instrument were processed with the ALCF (Section \ref{sec:alcf}) to again produce AVBC, cloud mask, cloud occurrence by height, and the total cloud fraction. The CHM 15k was used on four voyages (HMNZSW16, TAN1702, TAN1802, and NBP1704).

@@

Absolute calibration of the observed backscatter was performed by comparing the measured clear-sky molecular backscatter statistically with simulated clear-sky molecular backscatter. AVBC was resampled to 5 min temporal resolution and 50 m vertical resolution to increase the signal-to-noise ratio while having enough resolution to detect small-scale cloud variability. The noise standard deviation was calculated from AVBC at the highest range, where no clouds are expected. A cloud mask was calculated from AVBC using a fixed threshold of $\mathrm{2\times 10^{-6} m^{-1}sr^{-1}}$ after subtracting 5 standard deviations of range-scaled noise. Fig. \ref{fig:example}b shows an example of simulated Vaisala CL51 backscatter from ERA5 data, corresponding to a day of measurements by the instrument on the PS81/3 voyage.

@@

On the RV \emph{Polarstern} voyages, regular human-performed synoptic observations were available and included precipitation presence or absence and type. We used this dataset to train a convolutional artificial neural network (ANN) to recognize profiles with precipitation from lidar backscatter (Fig. \ref{fig:ann}a), implemented in the TensorFlow ANN framework \cite{tensorflow}. Samples of short time intervals (10 min) of near-surface lidar backscatter (0–250 m) were classified as clear, rain, snow, and fog, using the synoptic observations as a training dataset (Fig. \ref{fig:ann}b). From these, a binary, mutually exclusive classification of profiles as precipitating (rain or snow) or dry (clear or fog) was derived. For detecting model and reanalysis precipitation, we used a fixed threshold for surface precipitation flux of 0.1 mm h$^{-1}$ (the ANN was not used).

The ANN achieved 65\% sensitivity and 87\% specificity when the true positive rate (26\%) was made to match observations. The receiver operating characteristic curve is shown in Fig. \ref{fig:ann}c. We considered these rates satisfactory for the purpose of filtering precipitation profiles. Fig. \ref{fig:ann}d shows examples of the predicted precipitation compared to human-performed observations. The main ANN (`ANN` in Fig. \ref{fig:ann}) was trained on all data, and ancillary ANNs (`ANN2` in Fig. \ref{fig:ann}) were trained on all data except for each voyage. These were used for testing the results relative to each voyage.

@@

In addition to the above, we partitioned our data into two mutually exclusive subsets based on LTS, which is derived as the difference between the potential temperature at 700 hPa and the surface. Based on a histogram of LTS in ERA5 and MERRA-2 calculated at all voyage tracks and stations (Fig. \ref{fig:lts}), we determined a statistically-based dividing threshold of 12 K for weak stability ($<$ 12 K) and strong stability ($\geq$ 12 K) conditions.

@@

Fig. \ref{fig:cyclone-stability}a, b show a geographical distribution of the fraction of cyclonic days as determined by the cyclone tracking algorithm applied to the ERA5 reanalysis and ICON data (Section \ref{sec:cyclone-stability}). As expected, the strongest cyclonic activity is in the high-latitude SO zone, and it is relatively zonally symmetric at all latitudes. The pattern matches reasonably well with \citeA{hoskins2005}. While both reanalysis and the model agree relatively well, there are clear differences, particularly in the highest occurrence rate regions -- over the Amundsen Sea, where the Amundsen Sea low is located \cite{turner2013}, and which is more cyclonic in the reanalysis, and around Cape Adare, which is more cyclonic in ICON. These differences might, however, stem from the relatively short time periods of comparison (4 years) and the fact that the model is free-running.

Fig. \ref{fig:cyclone-stability}c, d show a geographical distribution of the conditions of weak and strong stability as determined by the LTS (Section \ref{sec:cyclone-stability}). Conditions of weak stability are prevalent in the mid-to-high SO (with respect to our SO partitioning; 50--65°S), which might be explained by the relatively cold near-surface air overlying the relatively warm sea surface. Conditions of strong stability are prevalent elsewhere over the SO. The distribution is also less zonally symmetric than the cyclonic activity. In the high-latitude SO, the presence of sea ice might have a substantial stabilising effect \cite{knight2024}. The ERA5 reanalysis is also substantially more stable than ICON across the whole region.

@@

We used the ALCF to derive cloud occurrence by height and the total cloud fraction from observations, ICON, ERA5, and MERRA-2. The results for all campaigns individually are shown in Fig. \ref{fig:cloud-occurrence-panel}. In addition, we aggregated the campaigns by calculating the averages and percentiles of all individual profiles, presented in Fig. \ref{fig:cloud-occurrence}. The analysis shows that the total cloud fraction (determined as the fraction of profiles with clouds at any height in the lidar cloud mask) is underestimated in ICON by about 10\% and in the reanalyses by about 20\%. When analyzed by height, ICON overestimates cloud occurrence below 1 km and underestimates it above; MERRA-2 underestimates cloud occurrence at all heights by up to about 10\%, especially near the surface; and ERA5 simulates cloud occurrence relatively well above 1 km but strongly underestimates it near the surface. We note that fog or near-surface clouds are strongly underestimated in the reanalyses (fog and clouds are both included in the cloud occurrence). As shown in Fig. \ref{fig:cloud-occurrence-panel}, the biases are relatively consistent across the campaigns and longitudes. We conclude that the ICON results the observations better than the reanalyses in this metric.

For all observations considered (Fig. \ref{fig:cloud-occurrence}a), the data show cloud occurrence peaking nearly at the surface, whereas the models show a higher peak (at about 500 m). The models generally underestimate the total cloud fraction by 10--20\% and show a strong reduction in cloud occurrence near the surface, which is not supported by the observations. ICON and ERA5 overestimate cloud occurrence at their peak (between 0--1 km). Above 1 km, ICON and MERRA-2 underestimate cloud occurrence, but ERA5 is accurate to about 3\% or less. The exaggerated peak in models is partly supported by the lifting condensation level (LCL) distribution, which peaks higher by about 200 m in the models than in the observations (nearly at the surface), although this is not very pronounced. This is indicative of near-surface relative humidity being often close to saturation in the observations but not in the models.

When subsetted by latitude (Fig. \ref{fig:cloud-occurrence}b, c), we see that the low-latitude SO zone displays a stronger peak of cloud occurrence near the surface than the high-latitude SO zone, and this could be because higher latitudes have less stable atmospheric profiles. The low- and high-latitude SO zones show similar biases in models as in the general case, but ERA5 does not overestimate the peak in the low-latitude SO zone (near-surface cloud occurrence is still strongly underestimated).

When subsetted by cyclonic and non-cyclonic situations (Fig. \ref{fig:cloud-occurrence}d, e), we see that the cyclonic situations have a larger amount of observed cloudiness, including peak and total cloud fraction, both by about 7\%. In the cyclonic situations, the model vertical profiles of cloud occurrence compare well with observations, but they peak higher by about 200 m and stronger by about 8\%. The reanalyses still tend to underestimate cloud occurrence above 1 km by about 5\% and near the surface by about 14\%. Non-cyclonic situations are similar to the general case, partially also because they form the majority of cases.

When subsetted by conditions of weak and strong stability (Fig. \ref{fig:cloud-occurrence}f, g), as defined in Section \ref{sec:cyclone-stability}, we see that in situations of strong stability, cloud occurrence peaks strongly nearly at the surface in observations, compared to situations of weak stability, where the peak is more diffuse between 0 and 1 km. It is worth mentioning that conditions of strong stability might be associated with the formation of advection fog, such as in situations of warm air advection from the north over colder sea surface thus inducing fog formation by cooling of the warm and humid air by the cold surface. In situations of strong stability, the models have smaller biases than in weak stability, with an overestimated peak by up to 12\%, underestimated cloud occurrence above 1 km by up to 5\%, and underestimated cloud occurrence near the surface by about 11\%. In situations of weak stability, the bias in ICON is very pronounced, with a much stronger peak at about 500 m; ERA5 underestimates cloud occurrence below 1 km (especially near the surface); and MERRA-2 underestimates cloud occurrence even more strongly.

@@

In Fig. \ref{fig:cloud-cover} we show the results for the same subsets of data as in Section \ref{sec:cloud-occurrence}. Observations display the highest proportion of high cloud cover values (5--8 oktas), peaking at 7 oktas. This pattern is not represented by ICON or either reanalysis. While ICON is closest to matching the observed distribution, it tends to be 1 okta clearer than the observations, peaking at 6 oktas, and substantially underestimating days with 8 oktas. Overall, the reanalyses show results similar to each other, underestimating cloud cover by about 2 oktas and strongly underestimating days with 7 and 8 oktas. Of the two reanalyses, MERRA-2 has slightly higher cloud cover than ERA5, by about 6\% at 6 octas, which makes it more consistent with observations.

When analyzed by subsets, observations in the cyclonic subset show the highest cloud cover, with 8 oktas occurring on one half of such days (Fig. \ref{fig:cloud-cover}d). This sensitivity to cyclonic conditions is not observed in ICON or the reanalyses. Interestingly, clear sky days (0 oktas) also have a local maximum peaking at about 15\% in this subset. When we contrast the low- and high-latitude zones, we see that the high-latitude zone tends to have greater cloud cover, peaking at 8 oktas (Fig. \ref{fig:cloud-cover}c). The high-latitude zone also has almost no clear sky or small cloud cover cases (0--4 oktas). ICON and the reanalyses represent this characteristic of the distribution well for 0--3 oktas, but otherwise show biases similar to the general case. One of the greatest biases is present in ERA5 in the subset of weak stability, in which ERA5 peaks at 3 oktas, while the observations peak at 7 oktas and show negligible cloud cover below 5 oktas.

\begin{figure}[p!]
\centering
\includegraphics[width=\textwidth]{img/clt_hist.pdf}
\caption{
Daily total cloud fraction histograms calculated as the average of all voyage and station histograms. The total cloud fraction of a day (UTC) is calculated as a fraction of cloudy (based on the cloud mask) observed (OBS) or simulated lidar profiles. The models and subsets are as in Fig. \ref{fig:cloud-occurrence}.
}
\label{fig:cloud-cover}
\end{figure}

\begin{figure}[p!]
\centering
\includegraphics[width=0.72\textwidth]{img/theta_hur.pdf}
\caption{
Virtual potential temperature ($\theta_v$) and relative humidity (RH) determined from radiosonde launches and co-located profiles in ICON, ERA5, and MERRA-2 in subsets as in Fig. \ref{fig:cloud-occurrence}. The solid lines are the average calculated from the averages of every individual voyage and station. The bands span the 16$^\mathrm{th}$--84$^\mathrm{th}$ percentiles calculated from the distribution of the voyage and station averages. Shown is also the relative frequency of occurrence and the number of profiles in each subset.
}
\label{fig:potential-temperature}
\end{figure}

@@

In order to link the cloud biases to the local physical conditions, we analyzed about 2300 radiosonde profiles south of 40°S from the 24 RV \emph{Polarstern} voyages, MARCUS, NBP1704, TAN1702, and TAN1802. Spatially and temporally colocated profiles were taken from ICON and the reanalyses. Because the time period covered by the ICON model output (2021--2024) was different from the time period covered by the observations (2010--2021), when comparing with the model, we first had to remap the observation time to model time by taking the same time relative to the start of the year. Consequently, we also had four virtual/model profiles (one for each year of 2021--2024) for each observed profile. The profiles were partitioned into the same subsets as above (Sections \ref{sec:cloud-occurrence} and \ref{sec:cloud-cover}). Apart from relative humidity, we focus on comparing virtual potential temperature ($\theta_v$) due to its role in low-level tropospheric stability, being one of the primary factors affecting shallow convection and the associated low-level cloud formation and dissipation. The observed and model profiles of virtual potential temperature are shown in Fig. \ref{fig:potential-temperature}.

Overall, the mean $\theta_v$ is accurate to less than 0.5 K in ICON and MERRA-2, being only colder by up to 2 K in the mid-to-high troposphere (less stable) in ICON than in observations (Fig. \ref{fig:potential-temperature}a). Larger differences exist, however, in the 40--55°S zone, where ICON is colder by about 5 K at higher altitudes (Fig. \ref{fig:potential-temperature}b). In other subsets, the bias is relatively small. MERRA-2 is very close to the observations, possibly due to a high accuracy of assimilation of this quantity. Notably, the variability of virtual potential temperature (as represented by the percentiles) is much smaller in ICON than in the observations. This indicates that the model's internal variability in the lower-tropospheric thermodynamic conditions in the SO is smaller than in reality.

Relative humidity displays much larger biases. In all subsets, ICON is too humid in the first 1 km by about 5\% but very accurate above, except for the 40--55°S zone and conditions of weak stability (Fig. \ref{fig:potential-temperature}b, g), where it is too dry between about 1 and 3 km. MERRA-2, on the other hand, is more humid than observations at all altitudes and in all subsets, by up to about 20\% at 5 km. Even though the mean near-surface relative humidity is similar to the observations (Fig. \ref{fig:potential-temperature}), the distribution in observation is more spread out across both high and low values, and thus observations have a greater prevalence of relative humidity close to 100\% and thus LCL located at the surface (Fig. \ref{fig:cloud-occurrence}a). In our calculations, LCL is an exclusive function of near-surface temperature, near-surface relative humidity, and surface pressure.

>>>>

Here, we conventionally refer to the SO as ocean regions south of 40°S, low-latitude SO as 40--55°S and high-latitude SO as south of 55°S. The reason for this dividing latitude is to split the SO into about two equal zones, as well as the results by \citeA{schuddeboom2021} (Fig.\ 2b), showing a contrast in CMIP model radiation biases, \citeA{schuddeboom2019} (Fig.\ 2) and \citeA{kuma2020} (Fig.\ 3) showing contrasting radiation biases in the Hadley Centre Global Environmental Model, and \citeA{cesana2022} showing contrasting cloud biases due to the 0°C isotherm reaching the surface at 55°S. The findings of \citeA{niu2024}, however, support a different dividing line of 62°S based on cloud condensation nuclei concentration.

@@

Together, we analyzed data from 31 voyages of RV \emph{Polarstern}, the resupply vessel (RSV) \emph{Aurora Australis}, RV \emph{Tangaroa}, RV \emph{Nathaniel B. Palmer}, Her (now His) Majesty's New Zealand Ship (HMNZS) \emph{Wellington} and one sub-Antarctic station (Macquarie Island) in the SO south of 40°S between 2010 and 2021. Fig.\ \ref{fig:map} shows a map of the campaigns, Table \ref{tab:voyages} lists the campaigns, and Table \ref{tab:voyage-references} lists references where available. Altogether, the analyzed dataset comprised 2421 days of data south of 40°S, but the availability of ceilometer data was slightly shorter due to gaps in measurements.

@@

The Vaisala CL51 and CT25K (photos in Fig.\ \ref{fig:map}d, e) are ceilometers operating at near-infrared wavelengths of 910 nm and 905 nm, respectively. The CL51 can also be configured to emulate the Vaisala CL31. The maximum range is 15.4 km (CL51), 7.7 km (CL31 emulation mode with 5 m vertical resolution), and 7.5 km (CT25K). The vertical resolution is 10 m (5 m configurable) in CL51 and 30 m in CT25K observations. The sampling (temporal) resolution is configurable, and in our datasets, it is approximately 6 s for CL51 on AA15‐16, 16 s for CT25K on MARCUS and MICRE, 36 s for CL51 on RV \emph{Polarstern}, and about 2.37 s for CL51 with CL31 emulation on TAN1502. The wavelengths of 905 and 910 nm are affected by water vapor absorption of about 20\% in the mid-latitudes \cite{wiegner2015,wiegner2019}, with 910 nm affected more strongly, but we do not expect this to be a significant issue as explained in \citeA{kuma2021}. The instrument data files containing raw uncalibrated backscatter were first converted to Network Common Data Form (NetCDF) with cl2nc (\url{https://github.com/peterkuma/cl2nc}) and then processed with the ALCF (Section \ref{sec:alcf}) to produce absolutely calibrated attenuated volume backscattering coefficient (AVBC), cloud mask, cloud occurrence by height, and the total cloud fraction. Because the CT25K uses a very similar wavelength to CL51, equivalent calculations as for CL51 were done assuming a wavelength of 910 nm. The Vaisala CL51 and CT25K instruments were used on most of the voyages and stations analyzed here. Fig.\ \ref{fig:example}a shows an example of AVBC derived from the CL51 instrument data.

@@

The Lufft CHM 15k (photo in Fig.\ \ref{fig:map}c) is a ceilometer operating at a near-infrared wavelength of 1064 nm. The maximum range is 15.4 km; the vertical resolution is 5 m in the near range (up to 150 m) and 15 m above; the sampling (temporal) resolution is 2 s; and the number of vertical levels is 1024. NetCDF files containing uncalibrated backscatter produced by the instrument were processed with the ALCF (Section \ref{sec:alcf}) to again produce AVBC, cloud mask, cloud occurrence by height, and the total cloud fraction. The CHM 15k was used on four voyages (HMNZSW16, TAN1702, TAN1802, and NBP1704).

@@

Absolute calibration of the observed backscatter was performed by comparing the measured clear-sky molecular backscatter statistically with simulated clear-sky molecular backscatter. AVBC was resampled to 5 min temporal resolution and 50 m vertical resolution to increase the signal-to-noise ratio while having enough resolution to detect small-scale cloud variability. The noise standard deviation was calculated from AVBC at the highest range, where no clouds are expected. A cloud mask was calculated from AVBC using a fixed threshold of $\mathrm{2\times 10^{-6} m^{-1}sr^{-1}}$ after subtracting 5 standard deviations of range-scaled noise. Fig.\ \ref{fig:example}b shows an example of simulated Vaisala CL51 backscatter from ERA5 data, corresponding to a day of measurements by the instrument on the PS81/3 voyage.

@@

On the RV \emph{Polarstern} voyages, regular human-performed synoptic observations were available and included precipitation presence or absence and type. We used this dataset to train a convolutional artificial neural network (ANN) to recognize profiles with precipitation from lidar backscatter (Fig.\ \ref{fig:ann}a), implemented in the TensorFlow ANN framework \cite{tensorflow}. Samples of short time intervals (10 min) of near-surface lidar backscatter (0–250 m) were classified as clear, rain, snow, and fog, using the synoptic observations as a training dataset (Fig.\ \ref{fig:ann}b). From these, a binary, mutually exclusive classification of profiles as precipitating (rain or snow) or dry (clear or fog) was derived. For detecting model and reanalysis precipitation, we used a fixed threshold for surface precipitation flux of 0.1 mm h$^{-1}$ (the ANN was not used).

The ANN achieved 65\% sensitivity and 87\% specificity when the true positive rate (26\%) was made to match observations. The receiver operating characteristic curve is shown in Fig.\ \ref{fig:ann}c. We considered these rates satisfactory for the purpose of filtering precipitation profiles. Fig.\ \ref{fig:ann}d shows examples of the predicted precipitation compared to human-performed observations. The main ANN (`ANN` in Fig.\ \ref{fig:ann}) was trained on all data, and ancillary ANNs (`ANN2` in Fig.\ \ref{fig:ann}) were trained on all data except for each voyage. These were used for testing the results relative to each voyage.

@@

In addition to the above, we partitioned our data into two mutually exclusive subsets based on LTS, which is derived as the difference between the potential temperature at 700 hPa and the surface. Based on a histogram of LTS in ERA5 and MERRA-2 calculated at all voyage tracks and stations (Fig.\ \ref{fig:lts}), we determined a statistically-based dividing threshold of 12 K for weak stability ($<$ 12 K) and strong stability ($\geq$ 12 K) conditions.

@@

Fig.\ \ref{fig:cyclone-stability}a, b show a geographical distribution of the fraction of cyclonic days as determined by the cyclone tracking algorithm applied to the ERA5 reanalysis and ICON data (Section \ref{sec:cyclone-stability}). As expected, the strongest cyclonic activity is in the high-latitude SO zone, and it is relatively zonally symmetric at all latitudes. The pattern matches reasonably well with \citeA{hoskins2005}. While both reanalysis and the model agree relatively well, there are clear differences, particularly in the highest occurrence rate regions -- over the Amundsen Sea, where the Amundsen Sea low is located \cite{turner2013}, and which is more cyclonic in the reanalysis, and around Cape Adare, which is more cyclonic in ICON. These differences might, however, stem from the relatively short time periods of comparison (4 years) and the fact that the model is free-running.

Fig.\ \ref{fig:cyclone-stability}c, d show a geographical distribution of the conditions of weak and strong stability as determined by the LTS (Section \ref{sec:cyclone-stability}). Conditions of weak stability are prevalent in the mid-to-high SO (with respect to our SO partitioning; 50--65°S), which might be explained by the relatively cold near-surface air overlying the relatively warm sea surface. Conditions of strong stability are prevalent elsewhere over the SO. The distribution is also less zonally symmetric than the cyclonic activity. In the high-latitude SO, the presence of sea ice might have a substantial stabilising effect \cite{knight2024}. The ERA5 reanalysis is also substantially more stable than ICON across the whole region.

@@

We used the ALCF to derive cloud occurrence by height and the total cloud fraction from observations, ICON, ERA5, and MERRA-2. The results for all campaigns individually are shown in Fig.\ \ref{fig:cloud-occurrence-panel}. In addition, we aggregated the campaigns by calculating the averages and percentiles of all individual profiles, presented in Fig.\ \ref{fig:cloud-occurrence}. The analysis shows that the total cloud fraction (determined as the fraction of profiles with clouds at any height in the lidar cloud mask) is underestimated in ICON by about 10\% and in the reanalyses by about 20\%. When analyzed by height, ICON overestimates cloud occurrence below 1 km and underestimates it above; MERRA-2 underestimates cloud occurrence at all heights by up to about 10\%, especially near the surface; and ERA5 simulates cloud occurrence relatively well above 1 km but strongly underestimates it near the surface. We note that fog or near-surface clouds are strongly underestimated in the reanalyses (fog and clouds are both included in the cloud occurrence). As shown in Fig.\ \ref{fig:cloud-occurrence-panel}, the biases are relatively consistent across the campaigns and longitudes. We conclude that the ICON results the observations better than the reanalyses in this metric.

For all observations considered (Fig.\ \ref{fig:cloud-occurrence}a), the data show cloud occurrence peaking nearly at the surface, whereas the models show a higher peak (at about 500 m). The models generally underestimate the total cloud fraction by 10--20\% and show a strong reduction in cloud occurrence near the surface, which is not supported by the observations. ICON and ERA5 overestimate cloud occurrence at their peak (between 0--1 km). Above 1 km, ICON and MERRA-2 underestimate cloud occurrence, but ERA5 is accurate to about 3\% or less. The exaggerated peak in models is partly supported by the lifting condensation level (LCL) distribution, which peaks higher by about 200 m in the models than in the observations (nearly at the surface), although this is not very pronounced. This is indicative of near-surface relative humidity being often close to saturation in the observations but not in the models.

When subsetted by latitude (Fig.\ \ref{fig:cloud-occurrence}b, c), we see that the low-latitude SO zone displays a stronger peak of cloud occurrence near the surface than the high-latitude SO zone, and this could be because higher latitudes have less stable atmospheric profiles. The low- and high-latitude SO zones show similar biases in models as in the general case, but ERA5 does not overestimate the peak in the low-latitude SO zone (near-surface cloud occurrence is still strongly underestimated).

When subsetted by cyclonic and non-cyclonic situations (Fig.\ \ref{fig:cloud-occurrence}d, e), we see that the cyclonic situations have a larger amount of observed cloudiness, including peak and total cloud fraction, both by about 7\%. In the cyclonic situations, the model vertical profiles of cloud occurrence compare well with observations, but they peak higher by about 200 m and stronger by about 8\%. The reanalyses still tend to underestimate cloud occurrence above 1 km by about 5\% and near the surface by about 14\%. Non-cyclonic situations are similar to the general case, partially also because they form the majority of cases.

When subsetted by conditions of weak and strong stability (Fig.\ \ref{fig:cloud-occurrence}f, g), as defined in Section \ref{sec:cyclone-stability}, we see that in situations of strong stability, cloud occurrence peaks strongly nearly at the surface in observations, compared to situations of weak stability, where the peak is more diffuse between 0 and 1 km. It is worth mentioning that conditions of strong stability might be associated with the formation of advection fog, such as in situations of warm air advection from the north over colder sea surface thus inducing fog formation by cooling of the warm and humid air by the cold surface. In situations of strong stability, the models have smaller biases than in weak stability, with an overestimated peak by up to 12\%, underestimated cloud occurrence above 1 km by up to 5\%, and underestimated cloud occurrence near the surface by about 11\%. In situations of weak stability, the bias in ICON is very pronounced, with a much stronger peak at about 500 m; ERA5 underestimates cloud occurrence below 1 km (especially near the surface); and MERRA-2 underestimates cloud occurrence even more strongly.

@@

In Fig.\ \ref{fig:cloud-cover} we show the results for the same subsets of data as in Section \ref{sec:cloud-occurrence}. Observations display the highest proportion of high cloud cover values (5--8 oktas), peaking at 7 oktas. This pattern is not represented by ICON or either reanalysis. While ICON is closest to matching the observed distribution, it tends to be 1 okta clearer than the observations, peaking at 6 oktas, and substantially underestimating days with 8 oktas. Overall, the reanalyses show results similar to each other, underestimating cloud cover by about 2 oktas and strongly underestimating days with 7 and 8 oktas. Of the two reanalyses, MERRA-2 has slightly higher cloud cover than ERA5, by about 6\% at 6 octas, which makes it more consistent with observations.

When analyzed by subsets, observations in the cyclonic subset show the highest cloud cover, with 8 oktas occurring on one half of such days (Fig.\ \ref{fig:cloud-cover}d). This sensitivity to cyclonic conditions is not observed in ICON or the reanalyses. Interestingly, clear sky days (0 oktas) also have a local maximum peaking at about 15\% in this subset. When we contrast the low- and high-latitude zones, we see that the high-latitude zone tends to have greater cloud cover, peaking at 8 oktas (Fig.\ \ref{fig:cloud-cover}c). The high-latitude zone also has almost no clear sky or small cloud cover cases (0--4 oktas). ICON and the reanalyses represent this characteristic of the distribution well for 0--3 oktas, but otherwise show biases similar to the general case. One of the greatest biases is present in ERA5 in the subset of weak stability, in which ERA5 peaks at 3 oktas, while the observations peak at 7 oktas and show negligible cloud cover below 5 oktas.

\begin{figure}[p!]
\centering
\includegraphics[width=\textwidth]{img/clt_hist.pdf}
\caption{
Daily total cloud fraction histograms calculated as the average of all voyage and station histograms. The total cloud fraction of a day (UTC) is calculated as a fraction of cloudy (based on the cloud mask) observed (OBS) or simulated lidar profiles. The models and subsets are as in Fig.\ \ref{fig:cloud-occurrence}.
}
\label{fig:cloud-cover}
\end{figure}

\begin{figure}[p!]
\centering
\includegraphics[width=0.72\textwidth]{img/theta_hur.pdf}
\caption{
Virtual potential temperature (virt.\ pot.\ temp.; $\theta_v$) and relative humidity (RH) determined from radiosonde launches and co-located profiles in ICON, ERA5, and MERRA-2 in subsets as in Fig.\ \ref{fig:cloud-occurrence}. The solid lines are the average calculated from the averages of every individual voyage and station. The bands span the 16$^\mathrm{th}$--84$^\mathrm{th}$ percentiles calculated from the distribution of the voyage and station averages. Shown is also the relative frequency of occurrence and the number of profiles in each subset.
}
\label{fig:potential-temperature}
\end{figure}

@@

In order to link the cloud biases to the local physical conditions, we analyzed about 2300 radiosonde profiles south of 40°S from the 24 RV \emph{Polarstern} voyages, MARCUS, NBP1704, TAN1702, and TAN1802. Spatially and temporally colocated profiles were taken from ICON and the reanalyses. Because the time period covered by the ICON model output (2021--2024) was different from the time period covered by the observations (2010--2021), when comparing with the model, we first had to remap the observation time to model time by taking the same time relative to the start of the year. Consequently, we also had four virtual/model profiles (one for each year of 2021--2024) for each observed profile. The profiles were partitioned into the same subsets as above (Sections \ref{sec:cloud-occurrence} and \ref{sec:cloud-cover}). Apart from relative humidity, we focus on comparing virtual potential temperature ($\theta_v$) due to its role in low-level tropospheric stability, being one of the primary factors affecting shallow convection and the associated low-level cloud formation and dissipation. The observed and model profiles of virtual potential temperature are shown in Fig.\ \ref{fig:potential-temperature}.

Overall, the mean $\theta_v$ is accurate to less than 0.5 K in ICON and MERRA-2, except for ICON being colder by up to 2.5 K in the mid-to-high troposphere (less stable) (Fig.\ \ref{fig:potential-temperature}a). Larger differences exist, however, in the 40--55°S zone, where ICON is colder by about 5 K at higher altitudes (Fig.\ \ref{fig:potential-temperature}b). In other subsets, the bias is relatively small. MERRA-2 is very close to the observations, possibly due to a high accuracy of assimilation of this quantity. Notably, the variability of virtual potential temperature (as represented by the percentiles) is much smaller in ICON than in the observations. This indicates that the model's internal variability in the lower-tropospheric thermodynamic conditions in the SO is smaller than in reality.

Relative humidity displays much larger biases. In all subsets, ICON is too humid in the first 1 km by about 5\% but very accurate above, except for the 40--55°S zone and conditions of weak stability (Fig.\ \ref{fig:potential-temperature}b, g), where it is too dry between about 1 and 3 km. MERRA-2, on the other hand, is more humid than observations at all altitudes and in all subsets, by up to about 20\% at 5 km. Even though the mean near-surface relative humidity is similar to the observations (Fig.\ \ref{fig:potential-temperature}), the distribution in observation is more spread out across both high and low values, and thus observations have a greater prevalence of relative humidity close to 100\% and thus LCL located at the surface (Fig.\ \ref{fig:cloud-occurrence}a). In our calculations, LCL is an exclusive function of near-surface temperature, near-surface relative humidity, and surface pressure.
